---
title: "Testing Differences"
format: 
  revealjs:
    incremental: true
editor: visual
embed-resources: true
---

## Variables

-   Dependent variable (a.k.a. $DV$, a.k.a. $Y$): The variable that *depends on* the independent variable

-   Independent variable (a.k.a. $IV$, a.k.a. $X$): The variable whose variation we're interested in

-   E.g., Does depression symptom severity (DV) *depend on* gender (IV)?

-   E.g., Do grades (DV) *depend on* depression symptom severity (IV)?

## Testing Differences

-   *t*-Test: Testing differences in means between 1-2 groups

-   ANOVA: Testing differences in means between 2+ groups

-   Chi-Square Test: Testing differences in proportions

## Getting Started

```{r}
#| echo: true
# Load packages
library(tidyverse)
library(here)

# Load data
hms_data <- readRDS(here("Healthy Minds Study", "Clean Data.rds"))

# Clean data
clean_data <- hms_data %>%
  drop_na()

# Describe data
str(clean_data)
```

## *t*-Test

A *t*-test tests differences between means. There are multiple versions of the *t*-test

-   One-sample *t*-test

-   Two-sample *t*-test

-   Paired-samples *t*-test

## One-Sample *t*-Test

One-sample *t*-test: Compares a mean to a pre-specified number (e.g., Does the average college student work more than 10 hours per week?)

-   $\mu_1 =$ mean hours worked per week

-   $h_0: \mu_1 \le 10$

-   $h_1: \mu_1 > 10$

## Two-Sample *t*-Test

Two-sample *t*-test: Compares two means from two different groups to one another (e.g., Does the average number of hours worked per week differ between male and female college students?)

-   $DV:$ hours worked per week

-   $IV:$ gender (male or female)

-   $\mu_1:$ mean hours worked per week among male students

-   $\mu_2:$ mean hours worked per week among female students

-   $h_0: \mu1 = \mu2$

-   $h_1: \mu_1 \ne \mu_2$

## Paired-Samples *t*-Test

Paired-samples *t*-test: Compares two means from the same group (e.g., Do college students experience less severe depression symptoms after they go through therapy?)

-   $DV:$ depression symptom severity

-   $IV:$ time-point (before or after therapy)

-   $\mu_1:$ mean depression symptom severity after therapy

-   $\mu_2:$ mean depression symptom severity before therapy

-   $h_0: \mu1 \ge \mu2$

-   $h_1: \mu_1 < \mu_2$

## Example: Two-Sample *t*-Test

RQ: Does the average amount of financial stress differ between male and female college students?

-   $DV:$ financial stress
-   $IV:$ gender (male or female)
-   $\mu_1:$ mean level of financial stress among male students
-   $\mu_2:$ mean level of financial stress among female students
-   $h_0: \mu_1 = \mu_2$
-   $h_1: \mu_1 \ne \mu_2$

## The Means

-   $\mu_1:$ mean level of financial stress among male students
-   $\mu_2:$ mean level of financial stress among female students

. . .

```{r}
#| echo: true
financial_stress_men <- clean_data$financial_stress[clean_data$gender == "Male"]
summary(financial_stress_men)
```

. . .

```{r}
#| echo: true
financial_stress_women <- clean_data$financial_stress[clean_data$gender == "Female"]
summary(financial_stress_women)
```

## The Distributions

Both sample distributions should be approximately normally distributed.

. . .

```{r}
#| echo: true
hist(financial_stress_men)
```

## The Distributions

Both sample distributions should be approximately normally distributed.

```{r}
#| echo: true
hist(financial_stress_women)
```

## The Analysis

We have our *sample* means. We have our *sample* distributions. What do we need to test whether these differences hold up in the *population*?

. . .

We need to know what we would see if we took *many different* samples. We need **sampling distributions**!

## Sampling Distribution of $\mu_1$

```{r}
#| echo: true
mean_financial_stress_men <- mean(financial_stress_men)
print(mean_financial_stress_men)
```

. . .

```{r}
#| echo: true
sd_financial_stress_men <- sd(financial_stress_men)
print(sd_financial_stress_men)
```

. . .

```{r}
#| echo: true
n_men <- length(financial_stress_men)
print(n_men)
```

. . .

```{r}
#| echo: true
se_financial_stress_men <- sd_financial_stress_men / sqrt(n_men)
print(se_financial_stress_men)
```

. . .

```{r}
#| echo: true
sampling_distribution_men <- rnorm(
  n = 100000, 
  mean = mean_financial_stress_men, 
  sd = se_financial_stress_men
)
```

## Sampling Distribution of $\mu_1$

```{r}
#| echo: true
tibble(sampling_distribution_men) %>%
  ggplot(aes(x = sampling_distribution_men)) +
  geom_density(binwidth = .001) +
  geom_vline(aes(xintercept = mean_financial_stress_men + (1.96 * se_financial_stress_men))) +
  geom_vline(aes(xintercept = mean_financial_stress_men - (1.96 * se_financial_stress_men)))
```

## Sampling Distribution of $\mu_2$

```{r}
#| echo: true
mean_financial_stress_women <- mean(financial_stress_women)
print(mean_financial_stress_women)
```

. . .

```{r}
#| echo: true
sd_financial_stress_women <- sd(financial_stress_women)
print(sd_financial_stress_women)
```

. . .

```{r}
#| echo: true
n_women <- length(financial_stress_women)
print(n_women)
```

. . .

```{r}
#| echo: true
se_financial_stress_women <- sd_financial_stress_women / sqrt(n_women)
print(se_financial_stress_women)
```

. . .

```{r}
#| echo: true
sampling_distribution_women <- rnorm(
  n = 100000, 
  mean = mean_financial_stress_women, 
  sd = se_financial_stress_women
)
```

## Sampling Distribution of $\mu_2$

```{r}
#| echo: true
tibble(sampling_distribution_women) %>%
  ggplot(aes(x = sampling_distribution_women)) +
  geom_density(binwidth = .001) +
  geom_vline(aes(xintercept = mean_financial_stress_women + (1.96 * se_financial_stress_women))) +
  geom_vline(aes(xintercept = mean_financial_stress_women - (1.96 * se_financial_stress_women)))
```

## Comparison of Sampling Distributions

```{r}
#| echo: true
tibble(
  sample_mean = c(sampling_distribution_men, sampling_distribution_women),
  x1 = c(rep("men", length(sampling_distribution_men)), rep("women", length(sampling_distribution_women)))
) %>%
  ggplot(aes(x = sample_mean, group = x1, color = x1, fill = x1)) +
  geom_density(binwidth = .001)
```

## The *t*-Statistic

We just reviewed the sampling distributions for our $\mu_1$ and $\mu_2$ statistics. The *t*-statistic represents the difference between the sample means, and has its own sampling distribution.

. . .

$t = \dfrac{\bar{x}_1 - \bar{x}_2}{se_{pooled}}$

. . .

$t=0$ when there is no difference in means

## The *t*-Statistic

```{r}
#| echo: true
difference_in_means <- mean_financial_stress_men - mean_financial_stress_women
print(difference_in_means)
```

. . .

```{r}
#| echo: true
se_pooled <- t.test(x = financial_stress_men, y = financial_stress_women)$stderr
print(se_pooled)
```

. . .

```{r}
#| echo: true
t <- difference_in_means / se_pooled
print(t)
```

## The *t*-Distribution Under $h_0$

```{r}
curve(dt(x, df = n_men + n_women - 1), 
      from = -4, 
      to = 4,
      main = "Sampling Distribution of t Under the Null Hypothesis",
      ylab = "Density",
      xlab = "t")
```

. . .

If the null hypothesis were true, what how likely would we be to see a *t*-statistic at least as extreme as -27.8?

## The Test

```{r}
#| echo: true
t.test(
  x = financial_stress_men, 
  y = financial_stress_women
)
```

## The Test (Again)

```{r}
#| echo: true
t.test(
  data = clean_data,
  financial_stress ~ gender,
  subset = gender %in% c("Male", "Female")
)
```

## Reporting the Test

In our sample, women experienced more financial stress (mean = 3.25, s.d. = 1.11) than men (mean = 3.00, s.d. = 1.13). We used a *t*-test to test whether this sample difference reflects a true difference in the population means. Our null hypothesis was that there would be no difference in population means. Our alternative hypothesis was that there would be some difference in population means. With a *p*-value \< .001, we reject the null hypothesis and conclude that there is a true difference in the population means.

## Visualizing the Difference

```{r}
clean_data %>%
  filter(gender %in% c("Male", "Female")) %>%
  group_by(gender) %>%
  summarize(mean_financial_stress = mean(financial_stress),
            sd_financial_stress = sd(financial_stress),
            se_financial_stress = sd(financial_stress) / sqrt(n())) %>%
  ggplot(aes(x = gender, y = mean_financial_stress)) +
  geom_col(color = "lightblue",
           fill = NA) +
  geom_errorbar(aes(ymin = mean_financial_stress - (se_financial_stress * 1.96),
                    ymax = mean_financial_stress + (se_financial_stress * 1.96)),
                width = .5) +
  theme_classic()
```

## ANOVA

ANOVA (short for "ANalysis Of VAriance") is used to test differences in means between 2+ groups

## Example: ANOVA

RQ: Does the average amount of financial stress depend on student gender (3 categories: male, female, and trans/gender diverse)?

-   $DV:$ financial stress
-   $IV:$ gender (male, female, or trans/gender diverse)
-   $\mu_1, \mu_2, \mu_3:$ mean level of financial stress among male, female, and trans/gender diverse students
-   $h_0: \mu_1 = \mu_2 = \mu_3$
-   $h_1: \mu_1 \ne \mu_2$ or $\mu_1 \ne \mu_3$ or $\mu_2 \ne \mu_3$

## The Means

-   $\mu_1:$ mean level of financial stress among male students
-   $\mu_2:$ mean level of financial stress among female students
-   $\mu_3:$ mean level of financial stress among trans/gender diverse students

. . .

```{r}
#| echo: true
financial_stress_tgd <- clean_data$financial_stress[clean_data$gender == "Trans/Gender-Diverse"]
summary(financial_stress_tgd)
```

. . .

```{r}
#| echo: true
summary(financial_stress_men)
```

. . .

```{r}
#| echo: true
summary(financial_stress_women)
```

## The *F*-Statistic

The "*F*-statistic" represents the ratio of variance *across* groups to the variance *within* groups. Like any statistic, it has its own sampling distribution.

. . .

$F = \dfrac{var_{across}}{var_{within}}$

. . .

$F=1$ when there is just as much variability within groups as there is across groups

## The *F*-Distribution Under $h_0$

```{r}
curve(df(x, df1 = 2, df2 = nrow(clean_data) - 3), 
      from = 0, 
      to = 5,
      main = "Sampling Distribution of F Under the Null Hypothesis",
      ylab = "Density",
      xlab = "F")
```

## The Test

```{r}
#| echo: true
financial_stress_by_gender <- aov(
  data = clean_data,
  formula = financial_stress ~ gender
)

summary(financial_stress_by_gender)
```

. . .

If the null hypothesis were true, what how likely would we be to see an *F*-statistic at least as large as 454.4?

## Reporting the Test

In our sample, trans and gender diverse students experienced the most financial stress (mean = 3.39, s.d. = 1.13), then women (mean = 3.25, s.d. = 1.11), then men (mean = 3.00, s.d. = 1.13). We used ANOVA to test whether these sample differences reflect a true differences in the population means. Our null hypothesis was that there would be no difference in population means. Our alternative hypothesis was that there would be some difference in population means. With a *p*-value \< .001, we reject the null hypothesis and conclude that there is a true difference in the population mean of financial stress by gender.

## Visualizing the Difference

```{r}
clean_data %>%
  group_by(gender) %>%
  summarize(mean_financial_stress = mean(financial_stress),
            sd_financial_stress = sd(financial_stress),
            se_financial_stress = sd(financial_stress) / sqrt(n())) %>%
  ggplot(aes(x = gender, y = mean_financial_stress)) +
  geom_col(color = "lightblue",
           fill = NA) +
  geom_errorbar(aes(ymin = mean_financial_stress - (se_financial_stress * 1.96),
                    ymax = mean_financial_stress + (se_financial_stress * 1.96)),
                width = .5) +
  theme_classic()
```

## Chi-Square Test

The Chi-Square test is used is used to test differences in proportions across any number of groups

## Example: Chi-Square Test

RQ: Does the frequency of anxiety depend on student's typical grades?

-   $DV:$ anxiety (present or absent)
-   $IV:$ student grades (A's, B's, etc.)
-   $h_0:$ the frequency of anxiety does not depend on a student's grades
-   $h_1$: the frequency of anxiety *does* depend on a student's grades

## The Table

```{r}
#| echo: true
anxiety_by_grades <- table(clean_data$anxiety, clean_data$grades)
print(anxiety_by_grades)
```

## The Table (Again)

```{r}
#| echo: true
clean_data %>%
  group_by(grades) %>%
  summarize(proportion_anxious = mean(anxiety))
```

The Chi-Square Statistic

The "chi square statistic" represents the difference between the observed frequency of our $DV$ at different levels of the $IV$, and what we would expect to see if there was no relationship between the two variables. Like any statistic, it has its own sampling distribution.

. . .

$\chi^2 = \sum\dfrac{(observed-expected)^2}{expected}$

. . .

$\chi^2$ is higher when there is a bigger difference between the observed and expected frequencies

## The *Chi-Square* Distribution Under $h_0$

```{r}
curve(dchisq(x, df = 6), 
      from = 0, 
      to = 20,
      main = "Sampling Distribution of Chi Square Under the Null Hypothesis",
      ylab = "Density",
      xlab = "Chi Square")
```

## The Test

```{r}
#| echo: true
chisq.test(anxiety_by_grades)
```

. . .

If the null hypothesis were true, what how likely would we be to see a chi square statistic at least as large as 640.8?

## Reporting the Test

In our sample, students with worse grades tended to experience anxiety more frequently. For example, 36% of students who typically received A's had clinical levels of anxiety symptoms, while 67% of students who typically receive F's did. We used a chi square test to test whether this sample difference reflects a true difference in the population proportions. Our null hypothesis was that there the frequency of anxiety does not depend on grades in the full population. Our alternative hypothesis was that the frequency of anxiety *does* depend on grades in the full population. With a *p*-value \< .001, we reject the null hypothesis and conclude that the frequency of anxiety really does depend on grades among the full population of college students.

## Visualizing the Difference

```{r}
clean_data %>%
  group_by(grades) %>%
  summarize(mean_anxiety = mean(anxiety),
            sd_anxiety = sd(anxiety),
            se_anxiety = sd(anxiety) / sqrt(n())) %>%
  ggplot(aes(x = grades, y = mean_anxiety)) +
  geom_col(color = "lightblue",
           fill = NA) +
  geom_errorbar(aes(ymin = mean_anxiety - (se_anxiety * 1.96),
                    ymax = mean_anxiety + (se_anxiety * 1.96)),
                width = .5) +
  theme_classic()
```
